### 2023-01-24 abstract.tex
First round of edits.

\begin{abstract}
\noindent
Sequence alignment is an essential method in bioinformatics and the basis of many analyses including phylogenetic inference, ancestral sequence reconstruction, and gene annotation. Errors made in alignment reconstruction can impact downstream analyses leading to erroneous conclusions in comparative and functional genomics.
Common artifacts in unpublished reference genomes include abiological frameshifts and early stop codons.
While for model organisms this is eventually fixed, for many species this is not the case and current curation efforts require discarding large amounts of data.
To address this need, we present COATi, a statistical codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data.
COATi will allow users to reduce the amount of discarded data while generating more accurate sequence alignments.
\end{abstract}

>>>>

\begin{abstract}
\noindent
Sequence alignment is an essential method in bioinformatics and the basis of many analyses, including phylogenetic inference, ancestral sequence reconstruction, and gene annotation.
Sequence artifacts and errors made in alignment reconstruction can impact downstream analyses leading to erroneous conclusions in comparative and functional genomic studies.
For example, abiological frameshifts and early stop codons are common artifacts found in protein coding sequences annotated in reference genomes.
While this is eventually fixed in the reference genomes of model organisms, many genomes contain theses artifacts, and researchers often discard large amounts of data in comparative genomic studies to prevent artifacts from impacting results.
To address this need, we present COATi, a statistical, codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data.
COATi will allow users to reduce the amount of discarded data while generating more accurate sequence alignments.
\end{abstract}

### 2023-01-24 introduction.tex
First round of edits.

Sequence alignment is a fundamental task in bioinformatics and a cornerstone step in comparative and functional genomic analysis \parencite{sequence_alignment_rosenberg_2009}.
Modern sequence analysis began with the heuristic homology algorithms of Needleman and Wunsch in 1970 \parencite{identification_smith_1981} and, while the methods developed since then have improved, alignment inference is not a solved problem \parencite{art_morrison_2015}.

A common strategy is to perform alignment inference in the amino acid space \parencite{bininda2005transalign,abascal2010translatorx}.
While this approach is an improvement over DNA models, it discards information, underperforms compared to alignment at the codon level, and fails in the presence of artifacts such as frameshifts and early stop codons.
Although some aligners incorporate codon substitution models, they do not support frameshifts or lack a statistical model.
In addition, while modeling indels to appear within codons is rare, this is often the case (Zhu \& Cartwright, personal communication, 2019).
Considering gaps to only appear between codons can result in missing the optimal alignment and inflate estimates of sequence divergence (Fig. \ref{fig:aln}).

>>>>

Sequence alignment is a fundamental task in bioinformatics and a cornerstone step in comparative and functional genomic analysis \parencite{sequence_alignment_rosenberg_2009}. While sophisticated advances have been made, the challenge of alignment inference has not been fully solved \parencite{art_morrison_2015}.
%Modern sequence analysis began with the heuristic homology algorithms of \textcite{Needleman1970} and \textcite{identification_smith_1981} and, while the methods developed since then have improved, alignment inference is not a solved problem \parencite{art_morrison_2015}. 
%
The alignment of protein coding DNA sequences is one such challenge, and a common approach to this problem is to perform alignment inference in amino-acid space \parencite[e.g.][]{bininda2005transalign,abascal2010translatorx}.
While this approach is an improvement over DNA models, it discards information, underperforms compared to alignment at the codon level, and fails in the presence of artifacts such as frameshifts and early stop codons.
Although some aligners incorporate codon substitution models, they do not support frameshifts or lack a statistical model.
In addition, while modeling indels to appear within codons is rare, this is often the case (\TODO{cite Ziqi's dissertation here and the mouse genome paper}).
Considering gaps to only appear between codons can result in missing the optimal alignment and inflate estimates of sequence divergence (Fig. \ref{fig:aln}).
